# Theme: Aquatic Ecosystems

**What:** Freshwater surface water temperature, oxygen, and chlorophyll-a.

**Where:** 7 lakes and 27 river/stream NEON sites.

**When:** Daily forecasts for at least 30-days in the future are accepted at any time. New forecast submissions are accepted daily. The only requirement is that submissions are predictions of the future at the time the forecast is submitted.

```{r echo = FALSE, message = FALSE}
library("tidyverse")
```

## Overview

In streams and rivers, forecasting water temperature can be meaningful for protecting aquatic communities while maintaining socio-economic benefits (Ouellet-Proulx et al. 2017). In lentic systems, successfully forecasting surface water temperatures can be important for fisheries and water utilities that need to manage the outflowing temperatures (Zhu et al. 2020). Recently, water temperature forecasts in lakes have been used to predict seasonal turnover when nutrients from the bottom can be mixed to the surface and impair the water quality.

Dissolved oxygen concentration is a critically important variable in limnology. Forecasts of dissolved oxygen in freshwaters is the first step to understanding other freshwater ecosystem processes. For example, oxygen serves as the gatekeeper to other biogeochemical reactions that occur in rivers and lakes. Preemptive forecasts of dissolved oxygen concentrations can anticipate periods of high or low oxygen availability, thereby providing insight into how the ecosystem may change at relatively short timescales.

chlorophyll-a is a metric of phytoplankton biomass. Phytoplankton biomass are the base of the aquatic food-web and an important indicator of water quality for managers.

## Challenge

This design challenge asks teams to produce forecasts of mean daily surface water temperature and/or dissolved oxygen in 7 NEON lake and/or 16 NEON river/stream sites for 30-days in the future. Additionally, forecasts of chlorophyll-a are invited for the 7 lakes and 3 non-wadeable river NEON sites.

You can chose to submit to either the lakes, rivers or streams or all three. You can also chose to submit any of the three focal variables (temperature, oxygen, and chlorophyll).

Teams are asked to submit their 30-day forecasts of NEON surface mean daily surface water temperature, dissolved oxygen, and/or chlorophyll-a along with uncertainty estimates and metadata. NEON surface water temperature, dissolved oxygen, and chlorophyll-a collected prior to the current date will be provided and may be used to build and improve the forecast models. Other data can be used as long as teams provide access (minimum of URL, but ideally a script) to all teams in the challenge.

## Data: Targets

The challenge uses the following NEON data products: - [DP1.20264.001](https://data.neonscience.org/data-products/DP1.20264.001){target="_blank"}: Temperature at specific depth in surface water for lakes - [DP1.20288.001](https://data.neonscience.org/data-products/DP1.20288.001){target="_blank"}: Water quality (includes oxygen and chlorophyll-a)\
- [DP1.20035.001](https://data.neonscience.org/data-products/DP1.20053.001){target ="\_blank"}: Temperature in surface waters for streams

Additionally, hourly water temperature profiles are also available for the lake sites, to aid model development, but will *not* be used in forecast scoring or evaluation. Only surface mean daily temperature forecasts should be submitted. Here is the format of the expanded targets data

### Surface Mean Daily Dissolved Oxygen Concentration

**Definition**

Dissolved oxygen (DO) is the concentration of oxygen dissolved in water. NEON's 30-minute time resolution from deployed water quality sondes among the freshwater sites reports this concentration as mg L<sup>-1</sup>. We have adapted the available NEON DO data to output the mean daily DO concentration in mg L<sup>-1</sup> from a water quality sonde(s) deployed in the top 1 m of the water column across all sites. Where multiple depths have observations within this depth range an average was taken. Common DO concentrations range between 0 and 12 mg L<sup>-1</sup> and DO concentrations less than 2 mg L<sup>-1</sup> are considered hypoxic.

**Motivation**

Dissolved oxygen concentration is a critically important variable in limnology. Forecasts of dissolved oxygen in freshwaters is the first step to understanding other freshwater ecosystem processes. For example, oxygen serves as the gatekeeper to other biogeochemical reactions that occur as well as determine the variety and health of aquatic organisms present in rivers and lakes. Preemptive forecasts of dissolved oxygen concentrations can anticipate periods of high or low oxygen availability, thereby providing insight into how the ecosystem may change at relatively short timescales.

### Surface Mean Daily Water Temperature

**Definition**

Water temperature is the temperature of the water. NEON's 30-minute time resolution from deployed water temperature sondes in the freshwater sites reports this in degrees Celsius (°C). We have adapted the available NEON water temperature data to output the mean daily water temperature in °C from temperature thermisters deployed 0-1m below the water surface at the lake sites and a water temperature sonde deployed in the stream sites. Where multiple depths have observations within this depth range an average was taken. Common water temperatures in lakes and streams range between 4 and 35 °C.

**Motivation**

In streams and rivers, forecasting water temperature can be meaningful for protecting aquatic communities while maintaining socio-economic benefits (Ouellet-Proulx et al. 2017). In lentic and lotic systems, successfully forecasting water temperatures can be important for management of fisheries and water utilities that rely on specific threshold temperatures (Zhu et al. 2020). Recently, lake temperature forecasts have been used to predict seasonal turnover, mixing bottom nutrients into the surface and impairing water quality.

### Chlorophyll-a

**Definition**

chlorophyll-a (chla) is the concentration of chlorophyll-a in the water column, as measured using florescence. NEON's 30-minute time resolution from deployed water quality sondes among the freshwater sites reports this concentration as mg L<sup>-1</sup>. We have adapted the available NEON chla data to output the mean daily chla concentration in mg L<sup>-1</sup> from a water quality sonde deployed in the top 1 m of the water column at a lake sites and water quality sondes deployed in the non-wadeable river sites (BLWA, FLNT, TOMB). No ongoing measurements of chla are available in the wadeable streams and therefore cannot be forecasted in this challenge.

**Motivation**

Phytoplankton biomass are the base of the aquatic food-web and an important indicator of water quality for managers.

### Focal sites

Information on the sites can be found here:

```{r message = FALSE}
site_data <- readr::read_csv("https://raw.githubusercontent.com/eco4cast/neon4cast-targets/main/NEON_Field_Site_Metadata_20220412.csv") |> 
  dplyr::filter(aquatics == 1)
```

```{r echo = FALSE}
site_data %>% 
  dplyr::select(field_site_id, field_site_name, field_site_subtype, field_latitude, field_longitude, neon_url) %>% 
  dplyr::rename(siteID = field_site_id,
         `site name` = field_site_name,
         `waterbody type` = field_site_subtype,
         `latitude` = field_latitude,
         `longtitude` =  field_longitude,
         `NEON site URL` = neon_url) %>% 
  dplyr::arrange(siteID) %>% 
  knitr::kable()
```

## Observed data latency

Through a collaboration with NEON, the new data will be available within 24-28 hrs of being collected for use in model training and forecast evaluation

## Partners

The challenge is hosted by the Ecological Forecasting Initiative (EFI; https://ecoforecast.org/) and its U.S. National Science Foundation sponsored Research Coordination Network (EFI-RCN; https://ecoforecast.org/rcn/).

Data used in the challenge are from the National Ecological Observatory Network (NEON): https://www.neonscience.org/.

Scientists from NOAA and USGS have been involved in the design of the challenge.

## References

Ouellet-Proulx, S., St-Hilaire, A., and Bouchar, M.-A.. 2017. Water temperature ensemble forecasts: Implementation using the CEQUEAU model on two contrasted river systems. Water 9(7):457. https://doi.org/10.3390/w9070457

Zhu, S., Ptak, M., Yaseen, Z.M., Dai, J. and Sivakumar, B. 2020. Forecasting surface water temperature in lakes: a comparison of approaches. Journal of Hydrology 585, 124809. https://doi.org/10.1016/j.jhydrol.2020.124809
