# Theme: Tick Populations

**What:** *Amblyomma americanum* nymphal tick abundance per sampled area

**Where:** 9 NEON sites

**When:** Forecasts for 4 weeks into the future using a weekly time-step

**Why:** There is a correlation between tick population abundance and disease incidence, meaning forecasts for tick abundance have the potential to aid in our understanding of disease risk through time and space.

**Who**: Open to any individual or team that registers

**How**: [REGISTER](https://nd.qualtrics.com/jfe/form/SV_9MJ29y2xNrBOjqZ){target="_blank"} your team and submit forecast. If you registered for the Round 1 (2021) and are using the same team and method then you do not need to re-register.

The video below is an overview of the Tick Populations Challenge that was recorded for the [2021 Early Career Annual Meeting](https://ecoforecast.org/ecological-forecasting-early-career-annual-meeting/){target="_blank"}

<iframe width="560" height="315" src="https://www.youtube.com/embed/lHH_nY52ZMM" title="YouTube video player" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen>

</iframe>

We held a Q&A session on March 24, 2021. You can find a recording from that session [HERE](https://youtu.be/kyR0MoDCO1Q){target="_blank"}.

```{r echo = FALSE, message = FALSE}
library("tidyverse")
```

## Overview

Target species for the population forecasts are *Amblyomma americanum* nymphal ticks. *A. americanum* is a vector of ehrlichiosis, tularemia, and southern tick-associated rash illness. The species is present in the eastern United States, and their populations are expanding. There is a correlation between tick population abundance and disease incidence, meaning forecasts for tick abundance have the potential to aid in our understanding of disease risk through time and space.

## Challenge

The challenge is open to any individual, group, or institution that may want to participate. The goals of this challenge are to forecast the density of *Amblyomma americanum* nymphs (ticks/1600m\^2) each epidemiological week (Sun-Sat) at nine NEON sites. Due to the latency in the taxonomic data reported by NEON, the challenge will be for the 2021 field season.

Teams must post information about any additional data they wish to use on the theme Slack channel so that other teams can potentially use the data as well.

## Data: Targets

The challenge uses the following NEON data products:

[DP1.10093.001](https://data.neonscience.org/data-products/DP1.10093.001){target="_blank"}: Ticks sampled using drag cloths

A file with previously released NEON data that has been processed into "targets" is provided below. The same processing will be applied to new data that are used for forecast evaluation. This [processing script](https://github.com/eco4cast/neon4cast-ticks/blob/master/02_ticks_targets.R){target="_blank"} is available in the [neon4cast-ticks](https://github.com/eco4cast/neon4cast-ticks){target="_blank"} GitHub repository.

### Amblyomma americanum nymphs

**Definition**

The density of *Amblyomma americanum* nymphs per week. Density is defined as the total number of individuals caught in a week across the forested plots divided by the total area sampled in the forested plots during the week. Densities are presented as ticks per 1600m\^2, as 1600m\^2 is the size of an individual NEON tick plot.

**Motivation**

We chose to use the density of *Amblyomma americanum* nymphs for several reasons. The first is that *Amblyomma americanum* is a vector of multiple pathogens, many of which cause human disease, and a forecast for their abundance could aid decisions in public health and personal protective measures. For simplicity, we chose to focus on one species for the abundance challenge, and the *Amblyomma americanum* nymphs are the most abundant tick observed in the NEON data. Most ticks are observed in to forested plots, and by standardizing the data to density of ticks observed per unit effort in the forested plots, we hope to avoid forecasters predicting sampling effort. We scaled the density to be representative of ticks per plot, which is more interpretable than ticks per square meter. Also, tick drags occur every three weeks. By having the challenge be for forecasting every week, participants won't have to predict which weeks drags occur.

### Focal sites

Information on the sites can be found here:

```{r message = FALSE}
site_data <- readr::read_csv("https://raw.githubusercontent.com/eco4cast/neon4cast-ticks/master/Ticks_NEON_Field_Site_Metadata_20210928.csv")
```

```{r echo = FALSE}
site_data %>% 
  select(field_site_id, field_site_name, field_dominant_nlcd_classes, field_latitude, field_longitude, neon_url) %>% 
  rename(siteID = field_site_id,
         `site name` = field_site_name,
         `vegetation type` = field_dominant_nlcd_classes,
         `latitude` = field_latitude,
         `longtitude` =  field_longitude,
         `NEON site URL` = neon_url) %>% 
  arrange(siteID) %>% 
  knitr::kable()
```

### Target data calculation

Tick drags occur every three weeks at the NEON sites used in this challenge. The sampling season at each site is determined by phenological milestones, beginning and ending within two weeks of green-up and senescence, respectively. The 1m\^2 cloth is dragged for 160m (and at least 80m), and ticks are collected intermittently. They are then sent to a lab for taxonomic identification. Ticks are then identified by life stage and taxonomic rank. The target data is for *Amblyomma americanum* nymphs that were identified to the species level; i.e. ticks identified as being in the *Amblyomma* genus are not included.

### Target file

Here is the format of the target file

```{r message = FALSE}
readr::read_csv("https://data.ecoforecast.org/neon4cast-targets/ticks/ticks-targets.csv.gz", guess_max = 1e6)
```

-   `time`: YYYY-MM-DD The first day (Sunday) of the MMWR week
-   `site_id`: Site where ticks are observed (HARV)
-   `variable`: `amblyomma_americanum` (density of *Amblyomma americanum* ticks; ticks / 1600m\^2)
-   `observed`: value for `variable`
-   `mmwr_week`: The MMWR week that starts on Sunday, consistent with CDC version of the epidemiological week\

## Timeline

Weekly-time step forecasts for a minimum of 4 weeks in the future submitted weekly by 12:59 pm ET on Mondays. A minimum of 4 week in the future must be forecasted for each submission, but they could be for longer. New forecasts can be submitted daily as new weather forecasts become available. The key is that submissions are predictions of the future.

Weekly submissions are allowed and encouraged as new observations and weather forecasts become available, therefore the automation of forecast generation may be ideal. There are many ways to automate scripts that are written to download observations and meteorology drivers, generate forecasts, and submit forecasts. Two tools that many have used are cron jobs (see the R package [cronR](https://cran.r-project.org/web/packages/cronR/index.html){target="_blank"}) that execute tasks at user specifics times and [github actions](https://youtu.be/dMrUlXi4_Bo){target="_blank"}. See more in @sec-faq

The timeline evaluation is determined by the data latency provided by NEON.  NEON currently lists 300 days as the latency between data collection and reporting data with taxonomic identification.  

## Submissions

The required names for forecasted variable: `amblyomma_americanum`

The required time unit: `date` for the first day (Sunday) of the MMWR week `date` in `YYYY-MM-DD` format.

Instructions for submitting forecasts are found in @sec-submissions

## Meterological inputs for modeling

Information about forecasted meteorology that is available for you to use when generating your forecasts can be found in @sec-met

## Useful functions

Functions for validating, evaluating and submitting forecasts can be found in @sec-functions

Functions for downloading and working with the meteorology forecasts can be be found in @sec-met

## Null models

The null model will be automatically generated each month, and is the historical mean for each week at each site. For weeks that don't have observations, the forecast is a linear interpolation between the preceding and following weeks.

## FAQ

Answers to frequency asks questions can be found in @sec-faq

## Design team

Matt Bitters, University of Colorado, Boulder\
Melissa Chen, University of Colorado, Boulder\
John Foster, Boston University\
Leah Johnson, Virginia Tech\
Shannon LaDeau, Cary Institute of Ecosystem Studies\
Cat Lippi, University of Florida\
Brett Melbourne, University of Colorado, Boulder\
Wynne Moss, University of Colorado, Boulder\
Sadie Ryan, University of Florida

## Partners

Data used in the challenge are collected by the National Ecological Observatory Network (NEON; https://www.neonscience.org/).
